|

### 2023-04-27 manuscript/manuscript_1.0.Rmd
New figure with simulations where learning pars evolve

---
title: |
  | The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones^[Laboratorio de Biología Evolutiva de Vertebrados, Departamento
  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia]
- C. Daniel Cadena*
- Redouan Bshary^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  # pdf_document:
  #   latex_engine: xelatex
fig_caption: yes
fontsize: 12pt
header-includes: "```{=latex}\n  \\usepackage{float}\n  \\usepackage{unicode-math}\n
  \ \\usepackage{setspace}\\doublespacing\n  \\usepackage[backref=page]{hyperref}\n
  \ \\linespread{2}\n  \\usepackage{lineno}\n  \\linenumbers\n  \\floatplacement{figure}{H}\n
  \ \\newcommand{\\beginsupplement}{ \\setcounter{table}{0}         \\renewcommand{\\thetable}{S\\arabic{table}}\\setcounter{figure}{0}
  \\renewcommand{\\thefigure}{S\\arabic{figure}}}\n```\n"
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: "Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while restraining
  from fighting individuals of higher quality. Communication systems involving signals
  of quality allow individuals to reduce uncertainty regarding the fighting ability
  of their partners and make more adaptive behavioral decisions. However, dishonest
  individuals can destabilize such communications systems. One open question is whether
  cognitive mechanisms, such as learning, can maintain the honesty of signals, thus,
  favoring their evolutionary stability. Here we present evolutionary simulations,
  where individuals can produce a signal proportional to their quality; and learn
  along their lifetime the best response to the signal emitted by their peers. In
  these simulations, learning on the receiver side can mediate the evolution of signals
  of quality on the sender side. When the cost of the signal is proportional to the
  quality of the sender populations are only composed of honest signalers. When the
  cost is not proportional to the quality of the signaler, the population is composed
  of both honest and dishonest signalers. We argue that learning can be a general
  cognitive mechanism playing a role in a wide range of communication systems.  \n"
---

@@

Individuals who survive engage in a series of pairwise interactions 
where they compete for resources. 
<!-- Individuals in real populations typically  -->
<!-- do not interact at random nor with all individuals in the population  -->
<!-- [@kurvers_Evolutionary_2014]. In order to allow for non-random interactions,  -->
<!-- we use the population vector to define an interaction neighbourhood  -->
<!-- for the focal individual, the size of the neighbourhood ($g$) -->
<!-- is the number of positions around the focal's from which the partner is drawn.  -->
<!-- So, if $g=N$ interactions are global; as $g$ gets smaller  -->
<!-- interactions are more local.  -->
In each interaction individuals must 
decide whether to escalate a fight or not. 
Following the classic *hawk-dove* game [@maynard-smith_Evolution_1982], 
if the focal individual fights and its partner does not, the focal gets 
as pay-off the resource of value $V$ while its partner gets nothing; 
if both individuals restrain  from fighting, then they split up the resource 
in half; if both decide to fight they split up the cost of fighting and 
the winner takes over the resource. We further assume that the 
probability of wining a fight for the focal individual depends on the 
difference in quality between it and its interacting partner; specifically 
it is given by $p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})$, where $k_3$ is a 
parameter defining how strong the quality difference determines the 
winning probability, and $i$ and $j$ denote the the position 
in the population vector of the focal and its interacting 
partner, respectively.

@@

@@

@@

@@

@@

We have presented here an evolutionary model where the response of a signal in a 
communication system is mediated by learning processes. Individuals 
throughout their life learn the best way to respond to a quantitative 
trait of their interactive partners. The learning process can mediate the evolution 
of an honest signal under the handicap 
principle [@grafen_Biological_1990;@zahavi_Mate_1975]. This conclusion is not 
different from the classical genetic models, where responses to the signal are
inherent to the individual. However, unlike classical genetic models,
our model shows that learning can also mediate a signal polymorphism in the 
absence of the handicap principle. Under this polymorphic equilibrium different 
types of individuals, with honest and dishonest signals, compose the population. 
The amount of information gathered through learning, as a well as the 
initial conditions of the learning process, are crucial for the emergence
and maintenance of this variation. The emergence of the polymorphism requires
a limit in the amount of information collected by individuals, as well as
peaceful behaviour in naive individuals.

@@

>>>>

---
title: |
  The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones^[Laboratorio de Biología Evolutiva de Vertebrados, Departamento
  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia]
- C. Daniel Cadena*
- Redouan Bshary^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex # default  # word_document: default
fig_caption: yes
fontsize: 12pt
header-includes: |
  ```{=latex}
    \usepackage{float}
    \usepackage{unicode-math}
    \usepackage{setspace}\doublespacing
    \usepackage[backref=page]{hyperref}
    \linespread{2}
    \usepackage{lineno}
    \linenumbers
    \floatplacement{figure}{H}
    \newcommand{\beginsupplement}{ \setcounter{table}{0}         \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: |
  Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while restraining
  from fighting individuals of higher quality. Communication systems involving signals
  of quality allow individuals to reduce uncertainty regarding the fighting ability
  of their partners and make more adaptive behavioral decisions. However, dishonest
  individuals can destabilize such communications systems. One open question is whether
  cognitive mechanisms, such as learning, can maintain the honesty of signals, thus,
  favoring their evolutionary stability. Here we present evolutionary simulations,
  where individuals can produce a signal proportional to their quality; and learn
  along their lifetime the best response to the signal emitted by their peers. In
  these simulations, learning on the receiver side can mediate the evolution of signals
  of quality on the sender side. When the cost of the signal is proportional to the
  quality of the sender populations are only composed of honest signalers. When the
  cost is not proportional to the quality of the signaler, the population is composed
  of both honest and dishonest signalers. We argue that learning can be a general
  cognitive mechanism playing a role in a wide range of communication systems."
---

@@

Individuals who survive engage in a series of pairwise interactions 
where they compete for resources. 
<!-- Individuals in real populations typically  -->
<!-- do not interact at random nor with all individuals in the population  -->
<!-- [@kurvers_Evolutionary_2014]. In order to allow for non-random interactions,  -->
<!-- we use the population vector to define an interaction neighbourhood  -->
<!-- for the focal individual, the size of the neighbourhood ($g$) -->
<!-- is the number of positions around the focal's from which the partner is drawn.  -->
<!-- So, if $g=N$ interactions are global; as $g$ gets smaller  -->
<!-- interactions are more local.  -->
In each interaction individuals must 
decide whether to escalate a fight or not. 
Following the classic *hawk-dove* game [@maynard-smith_Evolution_1982], 
if the focal individual fights and its partner does not, the focal gets 
as pay-off the resource of value $V$ while its partner gets nothing; 
if both individuals restrain  from fighting, then they split up the resource 
in half. If both individuals decide to fight, the winner takes over the resource and the cost of fighting is split among the two players. We further assume that the probability of wining a fight for the focal individual depends on the 
difference in quality between it and its interacting partner; specifically 
it is given by $p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})$, where $k_3$ is a 
parameter defining how strong the quality difference determines the 
winning probability, and $i$ and $j$ denote the the position 
in the population vector of the focal and its interacting 
partner, respectively.

@@

### Learning under a monomorphic population

@@

### Badges as Handicaps

@@

### The evolution of polymorphism mediated by learning

@@

rm(list=grep("temp",ls(),value = T))
```

### The peaceful, the clever and the aggressive

@@

Here we presented a model that integrates the basic mechanisms of 
learning with the fundamental rules of evolutionary processes. 
In the model the response of a signal in a 
communication system is mediated by learning processes. Individuals 
throughout their life learn the best way to respond to a quantitative 
trait of their interactive partners. 
The learning process can mediate the evolution 
of an honest signal under the handicap 
principle [@grafen_Biological_1990;@zahavi_Mate_1975]. This conclusion is not 
different from the classical genetic models, where responses to the signal are
inherent to the individual. However, unlike classical genetic models,
our model shows that learning can also mediate a signal polymorphism in the 
absence of the handicap principle. Under this polymorphic equilibrium different 
types of individuals, with honest and dishonest signals, compose the population. 
The amount of information gathered through learning, as a well as the 
initial conditions of the learning process, are crucial for the emergence
and maintenance of this variation. The emergence of the polymorphism requires
a limit in the amount of information collected by individuals, as well as
peaceful behaviour in naive individuals.

@@

```{r learnEvol,echo=FALSE,out.width='100%',fig.height=6,fig.align='center',fig.show='hold',fig.cap= "The evolution of costly signals. Portrait of the evolutionary dynamics of the sender code with snapsots of both sender and receiver codes just as in fig. \\ref{fig:handicap}. The middle panels show changes in the distribution of values for $\\alpha$ and $\\beta$ along evolutionary time. Panels above and below correspond to snapshots of the sender and receiver codes, respectively, generation time of the snapshots are indicated by the grey lines in the middle panels. Similarly to the evolutionary dynamics in Fig. \\ref{fig:handicap}, first the badge size evolves towards its minimum value. Then, the slope evolves positive values determining an increasing reaction norm. In contrast to \\ref{fig:handicap}, the normal distribution splits up into two modes. This translates into two types of reaction norms assorting in the population where both code for a positive relation between quality and badge size."}

scenario<-"betCostalphaL"#"alphaAct"#"nIntGroupNormQual"#
runChoi<-13

# Load files -------------------------------------------------------------------
# Project folder
listTest<-list.files(here("Simulations",paste0(scenario,"_")),full.names = TRUE)
evolList<-grep("evolLearn",listTest,value=TRUE)
indList<-grep("indLearn",listTest,value=TRUE)
paramName<-list.files(here("Simulations",paste0(scenario,"_")))
# paramName<-list.files(extSimsDir,full.names = TRUE)
paramName<-grep(".json",paramName,value=TRUE) %>% grep(pattern = "*0_*",value = TRUE)

param<-fromJSON(here("Simulations",paste0(scenario,"_"),paramName[1]))
  # fromJSON(paramName[1])

val<-1
fileId<-val

### Choose which interaction to visualize
# lastInt<-tail(pop[,unique(nInteract)],4)[2]
# # pop[,max(nInteract),by=.(seed,time)][,min(V1)]
# # to get last interaction: tail(pop[,unique(nInteract)],1)
# 
# popOneInd<-pop[nInteract==lastInt]
# vars<-c("alpha","beta","Badge")
# # ,"Badge",grep("WeightAct",names(popFinal),value = TRUE))
# popOneInd$idClust<-get_clusters(popOneInd,vars,k.max = 10,
#                                 Bsamples =500,iterMax = 500)
# clusSummary<-popOneInd[,.(meanAlph=mean(alpha),meanBet=mean(beta)),
#                        by=.(idClust,time,seed)]
# clusSummary[,orderClus:=0]
# invisible(capture.output(
#   for(i in 1:dim(clusSummary)[1]){
#   set(clusSummary,i,6L,match(clusSummary[i,meanAlph],
#         sort(clusSummary[time==clusSummary[i,time]&
#                            seed==clusSummary[i,seed]]$meanAlph)))
# }
# ))
# 
# 
# popOneInd<-merge(popOneInd,clusSummary,all = TRUE)
# 
# 
# fwrite(popOneInd,file = here("Simulations",paste0(scenario,"_"),
#                              paste0("popOneInd",nampar,Valpar,".txt")))
# 
popOneInd<-fread(here("Simulations",paste0(scenario,"_"),
                             paste0("popOneInd",nampar,Valpar,".txt")))

cols<-c("freqGenHawks","freqGenDove",  "freqGenEval",  #"freqGenLearn",
        "freqFenHawks", "freqFenDoves", "freqHH", "freqHD", "freqDD", "meanCue",
        "meanAlpha", "meanBeta",
        #"meanFit", #"meanInitCrit", "sdInitCrit", 
        "meanInitAct", "sdInitAct", "meanAlphaLearn","sdAlphaLearn",
        "WeightAct_0","WeightCrit_0",
        "WeightAct_1","WeightCrit_1","WeightAct_2",  "WeightCrit_2",
        "WeightAct_3",  "WeightCrit_3", "WeightAct_4",  "WeightCrit_4",
        "WeightAct_5","WeightCrit_5")

# get the trajectories for individual runs
traitsTrajs<-dcast(evol,time~seed,
                   value.var = c("meanAlpha","meanBeta",
                                 #"meanInitCrit",#"meanFit",
                                 "meanInitAct", "sdInitAct",
                                 "meanAlphaLearn","sdAlphaLearn",
                                 "sdAlpha","sdBeta","freqHH",
                                 "freqHD","freqDD"))

# Evolutionary dynamics for alpha - intercept of the sender --------------------------------------------
par(plt=posPlot(numploty = 3,idploty = 3,numplotx = 2,idplotx = 1,
                lowboundx = 8, title = TRUE,lowboundy = 10,
                upboundy = 92,upboundx = 92),
    mfrow=c(1,1), xaxt="s",las=1,mgp=c(2,0.8,0))

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend =
      expression(paste("intercept (",alpha[s],")")),col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("A.1"),cex=cexAxis,bty = "n",inset = -0.06)

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanAlpha_"),runChoi)],
         col=colGenesLin,lty = 1,type="l",lwd=5)

par(plt=posPlot(numploty = 3,idploty = 3,numplotx = 2, idplotx = 2,
                lowboundx = 8, title = TRUE,lowboundy = 10,
                upboundy = 92,upboundx = 92),
    xaxt="s",las=1,new=TRUE)

evolDist(indData = pop[seed==runChoi],variable = "beta",nbins = 10,pal = pal_dist,
         nlevels=10,cexAxis = 1.5,xlab="",ylab = "",xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,beta])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = expression(paste("Slope (",beta[s],")")),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("A.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,beta]))),
                      lty=1,col = "grey",lwd=2)

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanBeta_"),runChoi)],
         col=colGenesLin[2],lty = 1,type="l",lwd=4)

# Evolutionary dynamics for the speed of learning-------------------------------

par(plt=posPlot(numploty = 3,idploty = 2,numplotx = 2,idplotx = 1,
                lowboundx = 8, title = TRUE,lowboundy = 10,
                upboundy = 92,upboundx = 92),
    mfrow=c(1,1), xaxt="s",las=1,mgp=c(2,0.8,0),new=TRUE)

adjsYlims<-abs(range(pop[seed==runChoi,alphaLear])[1]-
                 range(pop[seed==runChoi,alphaLear])[2])*c(-0.1,0.2)

evolDist(indData = pop[seed==runChoi],variable = "alphaLear",nbins = 10,
         pal = pal_dist,
         nlevels=10,cexAxis = cexAxis,xlab="",ylab = "",xaxt="n",yaxt = "s",
         ylim = range(pop[seed==runChoi,alphaLear])+adjsYlims,
         xlim=c(0,22000))

mtext(text = "Trait value",cex = cexAxis,line = 1.8,side = 2,las=0)
legend("topright",legend = "Speed of learning",col=colGenesLin[1],lwd=2,
       bty = "n",cex=cexAxis)
legend("topleft",legend = c("B.1"),cex=cexAxis,bty = "n",inset = -0.06)

# black line showing the value 0 as a benchmark

lines(x = pop[,range(time)],y=c(0,0))

matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,alphaLear]),4),nrow = 2),
         lty=1,col = "grey",lwd=2)

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanAlphaLearn_"),runChoi)],         col=colGenesLin,lty = 1,type="l",lwd=5)

## Plot evolutionary dynamics of the naive behaviour of inds -------------------

par(plt=posPlot(numploty = 3,idploty = 2,numplotx = 2, idplotx = 2,
                lowboundx = 8, title = TRUE,lowboundy = 10,
                upboundy = 92,upboundx = 92),
    xaxt="s",las=1,new=TRUE)

adjsYlims<-abs(range(pop[seed==runChoi,initAct])[1]-
                 range(pop[seed==runChoi,initAct])[2])*c(-0.1,0.2)

evolDist(indData = pop[seed==runChoi],variable = "initAct",nbins = 10,
         pal = pal_dist, nlevels=10,cexAxis = 1.5,xlab="",ylab = "",
         xaxt="n",yaxt = "n",
         ylim = range(pop[seed==runChoi,initAct])+adjsYlims,xlim=c(0,22000))
legend("topright",legend = c("Behavioural tendency before learning"),cex=cexAxis,
       col=colGenesLin[2],lwd=2,bty = "n")
legend("topleft",legend = c("B.2"),cex=cexAxis,bty = "n",inset = -0.06)
# grey lines to show the generations shown in the upper and lower panels
matlines(x=matrix(rep(evolStats[genstoPrint,time],each=2),nrow=2),
         y=matrix(rep(range(pop[seed==runChoi,initAct]))),
                      lty=1,col = "grey",lwd=2)

matlines(x=traitsTrajs[,time],
         y=traitsTrajs[,.SD,
                       .SDcol=paste0(c("meanInitAct_"),runChoi)],
         col=colGenesLin[2],lty = 1,type="l",lwd=4)

# black line showing the value 0 as a benchmark

lines(x = pop[,range(time)],y=c(0,0))

# Choose time range
gen2plot<-round(seq(1,length(unique(evolStats$time)),length.out = 5))[2:5]
# Plor the actor
seqYax<-c("s",rep("n",3))
# seqYlabUp<-c("Badge",rep("",3))
seqYlabUp<-c("P(dove)",rep("",3))
seqYlabDown<-c("Badge",rep("",3))
seqXlabDown<-c("","Quality","")#paste0("seed: ",runChoi))
seqXlabUp<-c("","Badge","")
rangQual<-seq(0,1,length.out = 50)
interv<-1/(nCenters-1)
centers<-interv*seq(0,nCenters-1)
rangx<-seq(0,1,length=50)
count<-0
for(genC in genstoPrint){
  count<-count+1
  par(plt=posPlot(numplotx = 4,numploty = 3,idplotx = count,idploty = 1,
                  lowboundx = 8, title = TRUE,lowboundy = 10,
                  upboundy = 92,upboundx = 92),
      xaxt="s",las=1,new=TRUE)
  tempPop<-popOneInd[(time==unique(time)[genC]&seed==runChoi),.SD[.N],
               .SDcol=c(grep("WeightAct",
                             names(evol),value = TRUE),"Quality","alpha","beta",
                        "orderClus"),
               by=indId]

# meansClustmp<-tempPop[,.(alphaMean=mean(alpha),betaMean=mean(beta)),by=orderClus]
# reacNormpClust<-sapply(meansClustmp[,orderClus], function(x){
#   logist(rangx,alpha = meansClustmp[orderClus==x,alphaMean],
#          beta = meansClustmp[orderClus==x,betaMean])})
# matplot(x = rangx,y = reacNormpClust,col=colRuns[meansClustmp$orderClus],
#          type='l',cex.axis=cexAxis,
#          xlab="",ylab="",ylim=c(0,1.1),lty=1,
#          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  reacNormInd<-sapply(tempPop[indId %in%randInds,indId], function(x){
    logist(rangx,alpha = tempPop[indId==x,alpha],
           beta = tempPop[indId==x,beta])})
  matplot(x = rangx,y = reacNormInd,
          type='l',cex.axis=cexAxis,
          xlab="",ylab="",ylim=c(0,1.1),lty=1,col=1,
          lwd=2,yaxt=seqYax[count],xaxt="s",cex.lab=cexAxis)
  mtext(text = seqYlabDown[count],cex = cexAxis,line = 1.9,side = 2,las=0)
  mtext(text = seqXlabDown[count],cex = cexAxis,line = 1.7,side = 1,las=0)
  legend("topleft",legend = c(paste0("C.",count)),bty = "n",inset = -0.06)
  }

rm(list=grep("temp",ls(),value = T))
```
